---
title: "Tester Talk 26"
format: html
---
## Basic outline of process
Pull climatic gradient data
	- using rainfall and NDVI from Hawaii climate data portal from 3/11/2026
	  - rainfall metadata: 
	    Xresolution	0.00225
      Yresolution	0.00225
      ExtentXmin	-156.243
      ExtentXmax	-154.668
      ExtentYmin	18.849
      ExtentYmax	20.334
    - NDVI metadata: 
    Spatial resolution: 0.05° x 0.05° (about 5600 m, Climate Modeling Grid)
  - NDVI has much lower resolution than the rainfall data
Use formula to make probability surface for bird presence
	- Spectral gp (gaussian process simulation) 
Sample that bird surface for point and areal data
	- Use spatstat package in r for simulated point data with function rpoispp 
	- Areal data is split in to moku, which are further split into wao
	  - Moku data from [HI Statewide GIS program](https://geoportal.hawaii.gov/datasets/HiStateGIS::moku/about) 
	  - Wao are approximated by elevation
	    - elevation taken from [HI Statewide GIS program](https://geoportal.hawaii.gov/datasets/HiStateGIS::hawaii-elevation-contours-100ft/about) 
	    
	    
	    
Can hve multiple "trials" per ahapua'a, but overall modeled as avg probabiliyt of the slice, ran through some "noise" representing wwhether that observation was recorded

# setup

## Loading packages
```{r}
library(tidyverse)
library(here)
library(sf)
library(spectralGP)
library(spatstat)
library(terra)
library(INLA)
library(Matrix)
library(lwgeom)
library(fmesher)
library(spdep)
```

## Loading data
```{r}
moku <- st_read(here("data", "Tester26talk", "moku_cut.shp"))
rainfall <- rast(here("data", "Tester26talk", "HCDP_data", "rainfall_new_month_bi_data_map_2026_03.tif"))
ndvi <- rast(here("data", "Tester26talk", "HCDP_data", "ndvi_modis_day_bi_data_map_2026_03_01.tif"))
```

```{r}
#Take a quick look at the data
plot(st_geometry(moku))
plot(rainfall)
plot(ndvi)
```

# Generating and Cleaning data

## Cleaning
Doesn't look like rainfall and ndvi need to be changed at all, just the ahupuaa layer

```{r}
#| label: ahupuaa_clip

# clipping the extent of ahupuaa to the extent of the climatic data to remove the extra data :)

# make the coordinate reference systems match
ahupuaa <- moku |>
  st_transform(st_crs(rainfall)) |>
  st_make_valid()

# take bounding box of rainfall, intersect with the ahupaa layer
bbox <- st_bbox(rainfall) %>% 
  st_as_sfc()

ahupuaa_clip <- st_intersection(ahupuaa, bbox)

# check
plot(st_geometry(ahupuaa_clip)) # nice!
```

## Generate species distributions
Species distributed under gaussian process
- AKA they prefer a certain rainfall level and ndvi, with that preference level diminishing under a normal distr the further rainfall and ndvi values stray from their preferred level
Then apply noise to the data? 
Also, apply a gaussian random field?

```{r}
#| label: sp_prob_synthesis

# Rainfall ------------------------
# set the preferred climate for a species (akin to mean), as well as the standard deviation (akin to range of tolerance) arbitrarily lol 
pref_rainfall <- 600 # mean of dist
tolerance_rainfall <- 150 # sd of dist


# Apply a Gaussian response curve to raster
# P(occupancy) = exp(-2 * ((x - mean) / 2sd^2)^2) - pulled from gaussian curve fn, but removed first normalizing constant so it is always between 0-1
sp_rain_prob <- app(rainfall, function(x) {
  exp(-1 * (((x - pref_rainfall)^2) /((2*tolerance_rainfall)^2)))
})

names(sp_rain_prob) <- "occupancy_probability"


# plot to see 
plot(sp_rain_prob)

# NDVI -------------------------------------
# set the preferred climate for a species (mean), as well as the standard deviation (akin to range of tolerance) arbitrarily lol 
pref_ndvi <- .6 # mean
tolerance_ndvi <- .1 # sd

# Apply a Gaussian response curve to raster
# P(occupancy) = exp(-0.5 * ((x - optimum) / tolerance)^2) - pulled from gaussian curve fn, but removed first normalizing constant so it is always between 0-1
sp_ndvi_prob <- app(ndvi, function(x) {
  exp(-1 * (((x - pref_ndvi)^2) /((2*tolerance_ndvi)^2)))
})

names(sp_ndvi_prob) <- "occupancy_probability"


# plot to see 
plot(sp_ndvi_prob)

# combined probability ------------------------------------
prob_sp <- sp_ndvi_prob * sp_rain_prob
plot(prob_sp)
```


```{r}

# apply noise -----------------

## #Create a 2D GP object, generate noise 
# gridsize: should be 2^x
# using nrow() and ncol() to find resolution of raster
# for this raster it is 660X700 - 512 (2^9)should be ok to approx 1 node of variation per raster cell? 

gp_noise <- gp(gridsize = c(512, 512),
               specdens = matern.specdens,
               specdens.param = c(.1, 1), # c(smoothness, range), larger range = broader spatial correlation
               variance.param = 0.05) # height of the values p much - set here to .05 cuz it looks like it keeps the vals from -.5 to .5

simulate(gp_noise)  # modifies gp_noise in place, only returns null. don't need 2 put into new object

plot(gp_noise) # check it out! cool :)

# Extract the GP grid coordinates and values directly
gp_grid <- getgrid(gp_noise)     # returns list with $x1 and $x2 grid coords, checked they are from 0-1 as should be
gp_vals <- predict(gp_noise)     # returns values on the native GP grid

# gp_vals is a matrix of dimensions gridsize x gridsize (e.x. 512x512), the dim will be gridsize/2+1 (257) bc of fourier transformation
dim(gp_vals)

# Convert to raster, matching extent and CRS of sp prob
noise_rast <- rast(gp_vals, extent = ext(prob_sp), crs = crs(prob_sp))|>
  resample(prob_sp, method = "bilinear") # Resample to exactly match probability raster's resolution

plot(noise_rast) # ok nice noise field looks good 

# Center on 1 for multiplicative noise and apply to sp prob raster. Strength 
noise_centered <- noise_rast - global(noise_rast, mean)[[1]] + 1 # take each val, subtract the mean and add 1 
prob_noisy     <- prob_sp * noise_centered

# Plot to check that worked 
plot(prob_sp, main = "Original")
plot(prob_noisy,  main = "With spatialGP noise") # nice! but vals go way above 1, want to keep btwn 0 and 1 right?? 

# making the noisy sp dist normalized to be between 0 and 1
r_min <- minmax(prob_noisy)[1]
r_max <- minmax(prob_noisy)[2]
prob_noisy_norm <- (prob_noisy - r_min) / (r_max - r_min)

plot(prob_noisy_norm)
```


using the spectralGP mapping tool DIDNT WORK :( it would not properly map to the shape of the big island (first red flag) and had weird banding in it (second red flag) so instead I turned gp_grid into a raster and used terra's resample to fit it to the extent of the big island 

### Sample points from species dist

I will generate a set of coords to represent specimen collection. Then, I will clip these points to the extent of the simulated species range raster, then extract the sp occupancy probability as a new column on the points feature. Next, filter the points so that only points with a probability threshold >.25 are "detected" 
```{r}
#| label: point-sampling

# first generate the points, these will have coordinates on the unit square 
# width 6.6 and length 7 to make proportional to data bounds. intensity 6 makes 250 ish points

pp <- rpoispp(6, win = owin(c(0, 6.6), c(0,7)))
plot(pp) # looks good


# convert the units to real coordinates, make into new df
pp_coord_unit <- data.frame(x = xy2unit(pp[["x"]]), y = xy2unit(pp[["y"]]))

plot(pp_coord_unit) # ok nice converted to unit square

# Extract the real extent of your raster
ext_prob <- ext(prob_noisy_norm)  # gives xmin, xmax, ymin, ymax

xmin <- ext_prob$xmin
xmax <- ext_prob$xmax
ymin <- ext_prob$ymin
ymax <- ext_prob$ymax

# Extract the unit square coordinates
unit_coords <- cbind(pp_coord_unit$x, pp_coord_unit$y)

# Rescale to real coordinates
real_coords <- cbind(
  x = unit_coords[, 1] * (xmax - xmin) + xmin,
  y = unit_coords[, 2] * (ymax - ymin) + ymin
)

plot(real_coords) # ok looking good

# now clip points only to those on land ---------------------------

pts <- vect(as.data.frame(real_coords),
            geom = c("x", "y"),
            crs  = crs(prob_noisy_norm))

# Extract raster values at point locations
extracted <- terra::extract(prob_noisy_norm, pts)
# Keep only points that fall within the raster (non-NA values)
pts_clipped <- pts[!is.na(extracted[, 2]), ]
# i tried using join but it isn't working :( so I do this instead
extracted_clean <- drop_na(extracted, occupancy_probability)
# Add the extracted values as a new attribute column to the SpatVector
pts_clipped$occupancy_probability <- extracted_clean[ , 2]

# check the pts are in the right spot (over land only)
plot(prob_noisy_norm)
plot(pts_clipped, add = TRUE, col = "red", cex = 0.5) # nice! 

# remove some points based on their occupany probability -----------------
ggplot() + 
  geom_histogram(aes(x = pts_clipped$occupancy_probability)) # first check what the values might be - make sure the spread is ok - I want to have ~50 points remaining 
sample_pts <- pts_clipped[pts_clipped$occupancy_probability > 0.25, ]
sample_pts$occupancy_probability # nice! worked as intended
sample_pts$occupancy_probability <- NULL # now delete the prob value so we can make sure it is not going to affect our model

# check pts look like theyre only in high prob areas
length(sample_pts)
plot(prob_noisy_norm)
plot(sample_pts, add = TRUE, col = "red", cex = 0.5) 
```

### sample area from sp dist

```{r}
# --- Parameters ---
detection_threshold  <- 0.15    # mean occupancy > this = potential presence
knowledge_retention  <- 0.75    # probability knowledge is retained for presences
recent_extinction_threshold <- 0.3  # below this = likely recently extinct


prob_ahupuaa <- terra::extract(prob_noisy_norm, ahupuaa_clip, mean, bind = TRUE)

# Simulate TEK records

prob_ahupuaa$occupancy_probability_b <- sapply(1:nrow(prob_ahupuaa), function(i) {
  occ <- prob_ahupuaa$occupancy_probability[i]
  
  if (occ > detection_threshold) {
    # High occupancy = likely presence, BUT knowledge may still be lost
    knowledge_preserved <- rbinom(1, 1, knowledge_retention)
    if (knowledge_preserved) {
      return(1)   # presence recorded
    } else {
      return()  # keep the values for further processing
    }
    
  } else if (occ < recent_extinction_threshold) {
    # Low occupancy = likely recently extinct
    # Knowledge retained with probability knowledge_retention
    knowledge_preserved <- rbinom(1, 1, knowledge_retention)
    if (knowledge_preserved) {
      return(0)   # confirmed absence (knowledge retained of species absence)
    } else {
      return(NA)  # knowledge lost
    }
    
  } else {
    # Intermediate occupancy = uncertain either way
    return(NA)
  }
})

table(prob_ahupuaa$occupancy_probability_b, useNA = "always") # check that it worked

plot(prob_ahupuaa, "occupancy_probability") 
plot(prob_ahupuaa, "occupancy_probability_b") 
plot(st_geometry(ahupuaa_clip), add = TRUE) # looks ok? 
```


Alright, now I have the data I need to build the model. 

# Building the model

general structure of the model: 

Level 1 (Likelihood):
  Point observations  ~ Poisson point process(λ(s))   # intensity surface
  Polygon observations ~ Binomial(n, mean(p(s)))       # aggregated occupancy

Level 2 (Process):
  log(λ(s)) = β0 + β1 * rainfall(s) * ndvi(s) + w(s)           # log intensity
  logit(p(s)) = α0 + α1 * rainfall(s) * ndvi(s) + w(s)         # shared spatial effect

Level 3 (Priors):
  β0, β1, α0, α1 ~ Normal(0, 1)
  w(s) ~ GP(0, Matern covariance)                     # shared spatial random effect

## Preparing the data and generating mesh
```{r}
# Prepare data
# convert everything to planar crs -----------------------------
# certain functions only take planar coords
sample_pts_clean <- sample_pts |>
  project("EPSG:6628")

rainfall_clean <- rainfall |>
  project("EPSG:6628")

ndvi_clean <- ndvi |>
  project("EPSG:6628")

poly_sf <- prob_ahupuaa |>
  st_as_sf() |> # make sure it is an sf (sf::) rather than spatvector (terra::)
  st_transform("EPSG:6628")
```

### Point data 
```{r}
# point data ----------------------------------

# Extract rainfall and NDVI at presence point locations
pts_obs_df          <- as.data.frame(sample_pts_clean, geom = "XY")
pts_obs_df$rainfall <- extract(rainfall_clean, sample_pts_clean)[, 2]
pts_obs_df$ndvi <- extract(ndvi_clean, sample_pts_clean)[, 2]


```

### Area data

```{r}
# Subset to non-NA polygons for ZIP
# (ZIP uses both 0s and 1s - zeros are uncertain absences)
poly_sf     <- poly_sf[!is.na(poly_sf$occupancy_probability_b), ]
poly_sf$occupancy_probability <- poly_sf$occupancy_probability_b # make a new col 

table(poly_sf$occupancy_probability)   # check 0s and 1s - make sure no NA

# generating integration points for mesh
# Compute polygon areas ---
poly_sf$area <- as.numeric(st_area(poly_sf))

# Define scaling parameters 
min_points <- 5    # minimum points for smallest polygons
max_points <- 50   # maximum points for largest polygons

# Scale number of integration points proportionally to area
poly_sf$n_points <- round(
  min_points + (max_points - min_points) *
  (poly_sf$area - min(poly_sf$area)) /
  (max(poly_sf$area) - min(poly_sf$area))
)

# Check distribution of integration points
summary(poly_sf$n_points)
hist(poly_sf$n_points, main = "Integration points per polygon")

# Sample integration points proportional to area ---
poly_integration <- lapply(1:nrow(poly_sf), function(i) {
  pts <- st_sample(poly_sf[i, ],
                   size = poly_sf$n_points[i],
                   type = "regular")
  # Fallback to centroid if sampling fails
  if (length(pts) == 0) {
    pts <- st_centroid(poly_sf[i, ])
  }
  st_coordinates(pts)
})

# Verify points per polygon
sapply(poly_integration, nrow)
```

```{r}
# take integration pts and sample rainfall and ndvi 
# Average rainfall over each polygon
poly_sf$rainfall_avg <- sapply(poly_integration, function(pts) {
  pts_vect <- vect(as.data.frame(pts),
                   geom = c("X", "Y"),
                   crs  = crs(rainfall_clean))
  mean(extract(rainfall_clean, pts_vect)[, 2], na.rm = TRUE)
})

# Average ndvi over each polygon
poly_sf$ndvi_avg <- sapply(poly_integration, function(pts) {
  pts_vect <- vect(as.data.frame(pts),
                   geom = c("X", "Y"),
                   crs  = crs(rainfall_clean))
  mean(extract(ndvi_clean, pts_vect)[, 2], na.rm = TRUE)
})

```

### Mesh building
```{r}
# Prepare mesh
# Include presence points and polygon integration points in mesh
poly_all_coords <- do.call(rbind, poly_integration)

all_coords <- rbind(
  as.matrix(pts_obs_df[, c("x", "y")]),
  poly_all_coords
) # make into mnatrix

sum(duplicated(all_coords)) # make sure its 0 (no duplicates) to make sure mesh doesn't implode

# build mesh with starting coords as all_coords
mesh <- fm_mesh_2d_inla(
  loc      = all_coords, 
  crs = "EPSG:6628"
)

# Check mesh looks right
plot(mesh)
plot(st_geometry(poly_sf), add = TRUE, border = "red")
points(pts_obs_df$x, pts_obs_df$y, pch = 20, col = "blue", cex = 0.5)

# build SPDE with mesh --------
spde <- inla.spde2.matern(mesh  = mesh,
                           alpha = 2)

```


### Stacking data
#### point (LGCP) stack
```{r}
# log-gaussian cox process calc
# calculate  LGCP integration weights ---
mesh_coords_sf <- st_as_sf(as.data.frame(mesh$loc[, 1:2]),
                             coords = c("V1", "V2"))

# plot(mesh_coords_sf) look at mesh nodes - idk if it's supposed to be so unevenly dist? 

mesh_voronoi <- st_voronoi(st_union(mesh_coords_sf),
                            envelope = st_convex_hull(st_union(mesh_coords_sf)),
                           )
mesh_voronoi <- st_collection_extract(mesh_voronoi, "POLYGON")
mesh_voronoi <- st_make_valid(mesh_voronoi)

sf_use_s2(FALSE)

voronoi_centroids <- st_centroid(mesh_voronoi)
node_match        <- st_nearest_feature(mesh_coords_sf, voronoi_centroids)
mesh_voronoi      <- mesh_voronoi[node_match]

weights <- sapply(1:length(mesh_voronoi), function(i) {
  cell <- tryCatch(
    st_intersection(st_sfc(mesh_voronoi[i], crs = 4326), study_area),
    error = function(e) NULL
  )
  if (!is.null(cell) && length(cell) > 0) as.numeric(st_area(cell)) else 0
})

# make LGCP stack ----------------------------------
n_pts    <- nrow(pts_obs_df)
n_mesh   <- nrow(mesh$loc)

y_lgcp   <- c(rep(1, n_pts), rep(0, n_mesh))
e_lgcp   <- c(rep(0, n_pts), weights)

mesh_pts      <- vect(as.data.frame(mesh$loc[, 1:2]),
                       geom = c("V1", "V2"),
                       crs  = crs(rainfall_clean)) # use coord system of one of the layers (should all b same anyway)

# extract env variable values at mesh nodes
rainfall_mesh <- extract(rainfall_clean, mesh_pts)[, 2]
rainfall_mesh[is.na(rainfall_mesh)] <- mean(pts_obs_df$rainfall, na.rm = TRUE)

ndvi_mesh <- extract(ndvi_clean, mesh_pts)[, 2]
ndvi_mesh[is.na(ndvi_mesh)] <- mean(pts_obs_df$ndvi, na.rm = TRUE)

rainfall_lgcp <- c(pts_obs_df$rainfall, rainfall_mesh)
ndvi_lgcp <- c(pts_obs_df$ndvi, ndvi_mesh)

A_lgcp <- rbind(
  inla.spde.make.A(mesh, loc = as.matrix(pts_obs_df[, c("x", "y")])),
  Diagonal(n_mesh)
)

# make inla stack for the lgcp
stack_lgcp <- inla.stack(
  data    = list(y = y_lgcp,
                 e = e_lgcp),
  A       = list(A_lgcp, 1),
  effects = list(
    list(spatial   = 1:spde$n.spde),
    list(intercept = 1,
         rainfall  = rainfall_lgcp,
         ndvi      = ndvi_lgcp)      
  ),
  tag = "lgcp"
)
```

#### Polygon stack
```{r}
#| label: polygon_stack
# Polygon stack --------------------------------
A_poly_list <- lapply(poly_integration, function(pts) {
  if (nrow(pts) == 0) return(NULL)
  A_i <- inla.spde.make.A(mesh, loc = pts)
  Matrix::Matrix(Matrix::colMeans(A_i), nrow = 1)
})

# make sure everything is here and there aren't any nulls
valid_poly  <- !sapply(A_poly_list, is.null)
A_poly_avg  <- do.call(rbind, A_poly_list[valid_poly])
poly_sf_fit <- poly_sf[valid_poly, ]

stack_poly <- inla.stack(
  data    = list(y = poly_sf_fit$occupancy_probability,
                 e = rep(1, nrow(poly_sf_fit))),
  A       = list(A_poly_avg, 1),
  effects = list(
    list(spatial  = 1:spde$n.spde),
    list(intercept = 1,
         rainfall  = poly_sf_fit$rainfall_avg,
         ndvi      = poly_sf_fit$ndvi_avg)
  ),
  tag = "polygon"
)
```


#### Prediciton stack
```{r}
#| label: prediction_stack
# --- Build a prediction stack for the raster grid ---
raster_coords <- as.data.frame(prob_noisy_norm, xy = TRUE, na.rm = TRUE) # get raster coords, using rainfall just cuz
rainfall_pred      <- extract(rainfall,
                               vect(raster_coords,
                                    geom = c("x", "y"),
                                    crs  = crs(rainfall)))[, 2] # extract rainfall values at each coord
ndvi_pred          <- extract(ndvi,
                               vect(raster_coords,
                                    geom = c("x", "y"),
                                    crs  = crs(ndvi)))[, 2] # extract NDVI values at each coord

A_pred <- inla.spde.make.A(mesh,
                            loc = as.matrix(raster_coords[, c("x", "y")]))

stack_pred <- inla.stack(
  data    = list(y = rep(NA, nrow(raster_coords)),
                 e = rep(0,  nrow(raster_coords))),
  A       = list(A_pred, 1),
  effects = list(
    list(spatial   = 1:spde$n.spde),
    list(intercept = 1,
         rainfall  = rainfall_pred,
         ndvi      = ndvi_pred)
  ),
  tag = "pred"
)
```

#### Final stack of all three
```{r}
#| label: stack_all
# --- Rebuild full stack including prediction locations ---
stack_all  <- inla.stack(stack_lgcp, stack_poly, stack_pred)
stack_data <- inla.stack.data(stack_all)

# --- Recheck indices ---
index_lgcp <- inla.stack.index(stack_all, tag = "lgcp")$data
index_poly <- inla.stack.index(stack_all, tag = "polygon")$data
index_pred <- inla.stack.index(stack_all, tag = "pred")$data

cat("LGCP observations:", length(index_lgcp), "\n")
cat("Polygon observations:", length(index_poly), "\n")
cat("Prediction locations:", length(index_pred), "\n")  # should be 179685

# --- 4. Build response matrix ---
n_total  <- length(inla.stack.data(stack_all)$y)
y_matrix <- matrix(NA, nrow = n_total, ncol = 2)
colnames(y_matrix) <- c("y_lgcp", "y_poly")

y_matrix[index_lgcp, 1] <- y_lgcp
y_matrix[index_poly, 2] <- poly_sf_fit$occupancy_probability
```

## Model
```{r}
#| label: model

# the -1 is to remove the intercept, if not the model spits out a warning: "Warning in inla.core(formula = formula, family = family, contrasts = contrasts,  :  The A-matrix in the predictor (see ?control.predictor) is specified  but an intercept is in the formula. This will likely result  in the intercept being applied multiple times in the model, and is likely  not what you want. See ?inla.stack for more information. You can remove the intercept adding ``-1'' to the formula.

# the f() is for specifying spatial autocorrelation model

# Based on what I learned, this is what I think the formula should be
formula <- y ~ -1 + rainfall + ndvi + intercept + f(spatial, model = spde)

# Refit model with response matrix
model <- inla(formula,
              family            = c("poisson", "binomial"),
              data              = c(list(y = y_matrix), stack_data),
              E                 = stack_data$e,
              control.predictor = list(A       = inla.stack.A(stack_all),
                                       compute = TRUE),
              control.compute   = list(dic  = TRUE,
                                       waic = TRUE))

summary(model)
```


```{r}
#| label: plotting_mod_results
# Extract posterior mean and SD from linear predictor
# First get the index for prediction locations in the stack
index_pred <- inla.stack.index(stack_all, tag = "pred")$data

# Extract mean and SD of linear predictor at prediction locations
lp_mean <- model$summary.linear.predictor$mean[index_pred]
lp_sd   <- model$summary.linear.predictor$sd[index_pred]

summary(lp_sd)   # check these are non-zero

# Delta method approximation for SD on probability scale
# If p = plogis(eta), then dp/deta = p(1-p)
lp_mean_pred <- model$summary.linear.predictor$mean[index_pred]
lp_sd_pred   <- model$summary.linear.predictor$sd[index_pred]

p_mean <- plogis(lp_mean_pred)
p_sd   <- lp_sd_pred * p_mean * (1 - p_mean)   # delta method

# Add to raster coords and convert to raster
# Add to raster coords
raster_coords$pred_prob <- p_mean
raster_coords$pred_sd   <- p_sd

# Convert to rasters
prob_rast <- rast(raster_coords[, c("x", "y", "pred_prob")],
                  crs = crs(rainfall_clean))
sd_rast   <- rast(raster_coords[, c("x", "y", "pred_sd")],
                  crs = crs(rainfall_clean))

# Plot
par(mfrow = c(1, 2))
plot(prob_rast, main = "Predicted occupancy probability")
plot(sd_rast,   main = "Posterior uncertainty (SD)")
```


