---
title: "Tester Talk 26"
format: html
---
## Basic outline of process
Pull climatic gradient data
	- using rainfall and NDVI from Hawaii climate data portal from 3/11/2026
	  - rainfall metadata: 
	    Xresolution	0.00225
      Yresolution	0.00225
      ExtentXmin	-156.243
      ExtentXmax	-154.668
      ExtentYmin	18.849
      ExtentYmax	20.334
    - NDVI metadata: 
    Spatial resolution: 0.05° x 0.05° (about 5600 m, Climate Modeling Grid)
  - NDVI has much lower resolution than the rainfall data
Use formula to make probability surface for bird presence
	- Spectral gp (gaussian process simulation) 
Sample that bird surface for point and areal data
	- Use spatstat package in r for simulated point data with function rpoispp 
	- Areal data is split in to ahapua'a slices
	  - This data from [HI Statewide GIS program](https://geoportal.hawaii.gov/datasets/ahupuaa/explore?location=19.437569%2C-155.456033%2C9) 
	  
Can hve multiple "trials" per ahapua'a, but overall modeled as avg probabiliyt of the slice, ran through some "noise" representing wwhether that observation was recorded

# setup

## Loading packages
```{r}
library(tidyverse)
library(here)
library(sf)
library(spectralGP)
library(spatstat)
library(terra)
library(INLA)
library(Matrix)
library(lwgeom)
```

## Loading data
```{r}
ahupuaa_raw <- st_read(here("data", "Tester26talk", "Ahupuaa", "Ahupuaa.shp"))
rainfall <- rast(here("data", "Tester26talk", "HCDP_data", "rainfall_new_month_bi_data_map_2026_03.tif"))
ndvi <- rast(here("data", "Tester26talk", "HCDP_data", "ndvi_modis_day_bi_data_map_2026_03_01.tif"))
```

```{r}
#Take a quick look at the data
plot(st_geometry(ahupuaa_raw))
plot(rainfall)
plot(ndvi)
```

# Generating and Cleaning data

## Cleaning
Doesn't look like rainfall and ndvi need to be changed at all, just the ahupuaa layer

```{r}
#| label: ahupuaa_clip

# clipping the extent of ahupuaa to the extent of the climatic data to remove the extra data :)

# make the coordinate reference systems match
ahupuaa <- st_transform(ahupuaa_raw, st_crs(rainfall))

# take bounding box of rainfall, intersect with the ahupaa layer
bbox <- st_bbox(rainfall) %>% 
  st_as_sfc()

ahupuaa_clip <- st_intersection(ahupuaa, bbox)

# check
plot(st_geometry(ahupuaa_clip)) # nice!
```

## Generate species distributions
Species distributed under gaussian process
- AKA they prefer a certain rainfall level and ndvi, with that preference level diminishing under a normal distr the further rainfall and ndvi values stray from their preferred level
Then apply noise to the data? 
Also, apply a gaussian random field?

```{r}
#| label: sp_prob_synthesis

# Rainfall ------------------------
# set the preferred climate for a species (akin to mean), as well as the standard deviation (akin to range of tolerance) arbitrarily lol 
pref_rainfall <- 600
tolerance_rainfall <- 200


# Apply a Gaussian response curve to raster
# P(occupancy) = exp(-0.5 * ((x - optimum) / tolerance)^2) - pulled from gaussian curve fn, but removed first normalizing constant so it is always between 0-1
sp_rain_prob <- app(rainfall, function(x) {
  exp(-0.5 * ((x - pref_rainfall) / tolerance_rainfall)^2)
})

names(sp_rain_prob) <- "occupancy_probability"


# plot to see 
plot(sp_rain_prob)

# NDVI -------------------------------------
# set the preferred climate for a species (akin to mean), as well as the standard deviation (akin to range of tolerance) arbitrarily lol 
pref_ndvi <- .5
tolerance_ndvi <- .15

# Apply a Gaussian response curve to raster
# P(occupancy) = exp(-0.5 * ((x - optimum) / tolerance)^2) - pulled from gaussian curve fn, but removed first normalizing constant so it is always between 0-1
sp_ndvi_prob <- app(ndvi, function(x) {
  exp(-0.5 * ((x - pref_ndvi) / tolerance_ndvi)^2)
})

names(sp_ndvi_prob) <- "occupancy_probability"


# plot to see 
plot(sp_ndvi_prob)

# combined probability ------------------------------------
prob_sp <- sp_ndvi_prob * sp_rain_prob
plot(prob_sp)


# apply noise -----------------

## #Create a 2D GP object, generate noise 
# gridsize: should be 2^x
# using nrow() and ncol() to find resolution of raster
# for this raster it is 660X700 - 512 (2^9)should be ok to approx 1 node of variation per raster cell? 

gp_noise <- gp(gridsize = c(512, 512),
               specdens = matern.specdens,
               specdens.param = c(.2, 4), # c(smoothness, range), larger range = broader spatial correlation
               variance.param = 0.05) # height of the values p much - set here to .05 cuz it looks like it keeps the vals from -.5 to .5

simulate(gp_noise)  # modifies gp_noise in place, only returns null. don't need 2 put into new object

plot(gp_noise) # check it out! cool :)

# Extract the GP grid coordinates and process values directly
gp_grid <- getgrid(gp_noise)     # returns list with $x1 and $x2 grid coords, checked they are from 0-1 as should be
gp_vals <- predict(gp_noise)     # returns process values on the native GP grid

# gp_vals is a matrix of dimensions gridsize x gridsize (e.x. 512x512), the dim will be gridsize/2+1 (257) bc of fourier transformation
dim(gp_vals)

# Convert to raster, matching extent and CRS of sp prob
noise_rast <- rast(gp_vals, extent = ext(prob_sp), crs = crs(prob_sp))|>
  resample(prob_sp, method = "bilinear") # Resample to exactly match probability raster's resolution



plot(noise_rast) # ok nice noise field looks good 

# Center on 1 for multiplicative noise and apply to sp prob raster
noise_centered <- noise_rast - global(noise_rast, mean)[[1]] + 1 # take each val, subtract the mean and add 1 
prob_noisy     <- prob_sp * noise_centered

# Plot to check that worked 
plot(prob_sp, main = "Original")
plot(prob_noisy,  main = "With spatialGP noise") # nice! but vals go way above 1, want to keep btwn 0 and 1 right?? 

# making the noisy sp dist normalized to be between 0 and 1
r_min <- minmax(prob_noisy)[1]
r_max <- minmax(prob_noisy)[2]
prob_noisy_norm <- (prob_noisy - r_min) / (r_max - r_min)

plot(prob_noisy_norm)
```


using the spectralGP mapping tool DIDNT WORK :( it would not properly map to the shape of the big island (first red flag) and had weird banding in it (second red flag) so instead I turned gp_grid into a raster and used terra's resample to fit it to the extent of the big island 

## Sample points from species dist

I will generate a set of coords to represent specimen collection. Then, I will clip these points to the extent of the simulated species range raster, then extract the sp occupancy probability as a new column on the points feature. Next, filter the points so that only points with a probability threshold >.25 are "detected" 
```{r}
#| label: point-sampling

# first generate the points, these will have coordinates on the unit square 
# width 6.6 and length 7 to make proportional to data bounds. intensity 6 makes 250 ish points

pp <- rpoispp(7, win = owin(c(0, 6.6), c(0,7)))
plot(pp) # looks good


# convert the units to real coordinates, make into new df
pp_coord_unit <- data.frame(x = xy2unit(pp[["x"]]), y = xy2unit(pp[["y"]]))

plot(pp_coord_unit) # ok nice converted to unit square

# Extract the real extent of your raster
ext_prob <- ext(prob_sp)  # gives xmin, xmax, ymin, ymax

xmin <- ext_prob$xmin
xmax <- ext_prob$xmax
ymin <- ext_prob$ymin
ymax <- ext_prob$ymax

# Extract the unit square coordinates
unit_coords <- cbind(pp_coord_unit$x, pp_coord_unit$y)

# Rescale to real coordinates
real_coords <- cbind(
  x = unit_coords[, 1] * (xmax - xmin) + xmin,
  y = unit_coords[, 2] * (ymax - ymin) + ymin
)

plot(real_coords) # ok looking good

# now clip points only to those on land ---------------------------

pts <- vect(as.data.frame(real_coords),
            geom = c("x", "y"),
            crs  = crs(prob_noisy_norm))

# Extract raster values at point locations
extracted <- terra::extract(prob_noisy_norm, pts)
# Keep only points that fall within the raster (non-NA values)
pts_clipped <- pts[!is.na(extracted[, 2]), ]
# i tried using join but it isn't working :( so I do this instead
extracted_clean <- drop_na(extracted, occupancy_probability)
# Add the extracted values as a new attribute column to the SpatVector
pts_clipped$occupancy_probability <- extracted_clean[ , 2]

# check the pts are in the right spot (over land only)
plot(prob_noisy_norm)
plot(pts_clipped, add = TRUE, col = "red", cex = 0.5) # nice! 

# remove some points based on their occupany probability -----------------
pts_clipped$occupancy_probability # first check what the values might be - make sure the spread is good 
sample_pts <- pts_clipped[pts_clipped$occupancy_probability > 0.25, ]
sample_pts$occupancy_probability # nice! worked as intended
sample_pts$occupancy_probability <- NULL # now delete the prob value so we can make sure it is not going to affect our model

# check pts look like theyre only in high prob areas
length(sample_pts)
plot(prob_noisy_norm)
plot(sample_pts, add = TRUE, col = "red", cex = 0.5) 
```

## sample area from sp dist
```{r}
#| label: areal-sampling

plot(st_geometry(ahupuaa_clip)) # there's a lot of rly big ahupuaas, maybe I should cut up by wao

prob_ahupuaa <- terra::extract(prob_noisy_norm, ahupuaa_clip, mean, bind = TRUE) # extract mean occ prob for each ahupuaa

prob_ahupuaa$occupancy_probability <- ifelse(prob_ahupuaa$occupancy_probability > 0.1, 1, 0) # then make it so if the mean prob is >.1, then it has been observed here. otherwise not. 

length(prob_ahupuaa$occupancy_probability)

```

Alright, now I have the data I need to build the model. 

# Building the model

general structure of the model: 

Level 1 (Likelihood):
  Point observations  ~ Poisson point process(λ(s))   # intensity surface
  Polygon observations ~ Binomial(n, mean(p(s)))       # aggregated occupancy

Level 2 (Process):
  log(λ(s)) = β0 + β1 * rainfall(s) * ndvi(s) + w(s)           # log intensity
  logit(p(s)) = α0 + α1 * rainfall(s) * ndvi(s) + w(s)         # shared spatial effect

Level 3 (Priors):
  β0, β1, α0, α1 ~ Normal(0, 1)
  w(s) ~ GP(0, Matern covariance)                     # shared spatial random effect


```{r}
# Prepare data

# point data ----------------------------------
# Extract rainfall and NDVI at presence point locations
pts_obs_df          <- as.data.frame(sample_pts, geom = "XY")
pts_obs_df$rainfall <- extract(rainfall, sample_pts)[, 2]
pts_obs_df$ndvi <- extract(ndvi, sample_pts)[, 2]

# Prepare polygon data ------------------------

poly_sf <- st_as_sf(prob_ahupuaa) # make sure it is an sf rather than spatvector

# Sample integration points within each polygon and average rainfall
n_integration <- 25   # adjust based on polygon size and desired accuracy

poly_integration <- lapply(1:nrow(poly_sf), function(i) {
  pts <- st_sample(poly_sf[i, ], size = n_integration, type = "regular")
  # st_sample may return fewer points than requested for small polygons
  # fall back to centroid if no points returned
  if (length(pts) == 0) {
    pts <- st_centroid(poly_sf[i, ])
  }
  st_coordinates(pts)
})

# Average rainfall covariate over each polygon
poly_sf$rainfall_avg <- sapply(poly_integration, function(pts) {
  pts_vect <- vect(as.data.frame(pts),
                   geom = c("X", "Y"),
                   crs  = crs(rainfall))
  mean(extract(rainfall, pts_vect)[, 2], na.rm = TRUE)
})

# Average ndvi covariate over each polygon
poly_sf$ndvi_avg <- sapply(poly_integration, function(pts) {
  pts_vect <- vect(as.data.frame(pts),
                   geom = c("X", "Y"),
                   crs  = crs(rainfall))
  mean(extract(ndvi, pts_vect)[, 2], na.rm = TRUE)
})
```


```{r}
# Prepare mesh
# Include presence points and polygon integration points in mesh
poly_all_coords <- do.call(rbind, poly_integration)

all_coords <- rbind(
  as.matrix(pts_obs_df[, c("x", "y")]),
  poly_all_coords
) # make into mnatrix

sum(duplicated(all_coords)) # Check all_coords for duplicates to make sure mesh doesn't implode

# Rebuild mesh with explicit boundary and coordinates in degrees
# max.edge should be in degrees for WGS84 — 0.1 degrees ~ 11km
mesh <- inla.mesh.2d(
  loc      = all_coords,
  max.edge = c(0.05, 0.2),   # inner and outer edge length in degrees
  cutoff   = 0.01,            # minimum distance between nodes in degrees
  offset   = c(0.1, 0.3)     # inner and outer boundary extension in degrees
)

# Check mesh looks right
plot(mesh)
plot(st_geometry(poly_sf), add = TRUE, border = "red")
points(pts_obs_df$x, pts_obs_df$y, pch = 20, col = "blue", cex = 0.5)

# build SPDE with mesh --------
spde <- inla.spde2.matern(mesh  = mesh,
                           alpha = 2)

```

```{r}
# log-gaussian cox process calc
# calculate  LGCP integration weights ---
mesh_coords_sf <- st_as_sf(as.data.frame(mesh$loc[, 1:2]),
                             coords = c("V1", "V2"),
                             crs    = 4326)

mesh_voronoi <- st_voronoi(st_union(mesh_coords_sf),
                            envelope = st_convex_hull(st_union(mesh_coords_sf)))
mesh_voronoi <- st_collection_extract(mesh_voronoi, "POLYGON")
mesh_voronoi <- st_make_valid(mesh_voronoi)

sf_use_s2(FALSE)

voronoi_centroids <- st_centroid(mesh_voronoi)
node_match        <- st_nearest_feature(mesh_coords_sf, voronoi_centroids)
mesh_voronoi      <- mesh_voronoi[node_match]

weights <- sapply(1:length(mesh_voronoi), function(i) {
  cell <- tryCatch(
    st_intersection(st_sfc(mesh_voronoi[i], crs = 4326), study_area),
    error = function(e) NULL
  )
  if (!is.null(cell) && length(cell) > 0) as.numeric(st_area(cell)) else 0
})

# make LGCP stack ----------------------------------
n_pts    <- nrow(pts_obs_df)
n_mesh   <- nrow(mesh$loc)

y_lgcp   <- c(rep(1, n_pts), rep(0, n_mesh))
e_lgcp   <- c(rep(0, n_pts), weights)

mesh_pts      <- vect(as.data.frame(mesh$loc[, 1:2]),
                       geom = c("V1", "V2"),
                       crs  = crs(rainfall)) # use coord system of one of the layers (should all b same anyway)

# extract env variable values at mesh nodes
rainfall_mesh <- extract(rainfall, mesh_pts)[, 2]
rainfall_mesh[is.na(rainfall_mesh)] <- mean(pts_obs_df$rainfall, na.rm = TRUE)

ndvi_mesh <- extract(ndvi, mesh_pts)[, 2]
ndvi_mesh[is.na(ndvi_mesh)] <- mean(pts_obs_df$ndvi, na.rm = TRUE)

rainfall_lgcp <- c(pts_obs_df$rainfall, rainfall_mesh)
ndvi_lgcp <- c(pts_obs_df$ndvi, ndvi_mesh)

A_lgcp <- rbind(
  inla.spde.make.A(mesh, loc = as.matrix(pts_obs_df[, c("x", "y")])),
  Diagonal(n_mesh)
)

# make inla stack 
stack_lgcp <- inla.stack(
  data    = list(y = y_lgcp,
                 e = e_lgcp),
  A       = list(A_lgcp, 1),
  effects = list(
    list(spatial   = 1:spde$n.spde),
    list(intercept = 1,
         rainfall  = rainfall_lgcp,
         ndvi      = ndvi_lgcp)      
  ),
  tag = "lgcp"
)

# Polygon stack --------------------------------
A_poly_list <- lapply(poly_integration, function(pts) {
  if (nrow(pts) == 0) return(NULL)
  A_i <- inla.spde.make.A(mesh, loc = pts)
  Matrix::Matrix(Matrix::colMeans(A_i), nrow = 1)
})

# make sure everything is here and there aren't any nulls
valid_poly  <- !sapply(A_poly_list, is.null)
A_poly_avg  <- do.call(rbind, A_poly_list[valid_poly])
poly_sf_fit <- poly_sf[valid_poly, ]

stack_poly <- inla.stack(
  data    = list(y = poly_sf_fit$occupancy_probability,
                 e = rep(1, nrow(poly_sf_fit))),
  A       = list(A_poly_avg, 1),
  effects = list(
    list(spatial  = 1:spde$n.spde),
    list(intercept = 1,
         rainfall  = poly_sf_fit$rainfall_avg,
         ndvi      = poly_sf_fit$ndvi_avg)    
  ),
  tag = "polygon"
)
```


```{r}
# Get the combined data from the stack
stack_all <- inla.stack(stack_lgcp, stack_poly)
stack_data <- inla.stack.data(stack_all)

# Total number of observations
n_total <- length(stack_data$y)
n_lgcp  <- length(y_lgcp)
n_poly  <- nrow(poly_sf_fit)

# Create response matrix with one column per likelihood
# NAs tell INLA to ignore that observation for that likelihood
y_matrix <- matrix(NA, nrow = n_total, ncol = 2)
colnames(y_matrix) <- c("y_lgcp", "y_poly")

# Fill in LGCP responses (first rows)
y_matrix[1:n_lgcp, 1] <- y_lgcp

# Fill in polygon responses (remaining rows)
y_matrix[(n_lgcp + 1):n_total, 2] <- stack_data$y[(n_lgcp + 1):n_total]

stack_all <- inla.stack(stack_lgcp, stack_poly)
```


```{r}
#| label: model

# this is what claude gave me - I don't agree 
#formula <- y ~ -1 + intercept + rainfall + ndvi + f(spatial, model = spde) 
# why does it have a -1 intercept? why is intercept a predictor?? the f() is for choosing the random effect, i think??

# Based on what I learned, this is what I think the formula should be
formula <- y ~ rainfall + ndvi + f(____, model = "besag") + f(_____, model = "iid") 
# rainfall and ndvi are predictors, f( model = "besag") is for spatial random effect and f( model = "iid") is for the unstructured random effect

# Refit model with response matrix
model <- inla(formula,
              family            = c("poisson", "binomial"),
              data              = c(list(y = y_matrix), stack_data),
              E                 = stack_data$e,
              control.predictor = list(A       = inla.stack.A(stack_all),
                                       compute = TRUE),
              control.compute   = list(dic  = TRUE,
                                       waic = TRUE))

summary(model)
```


```{r}
 
# Project predictions back to raster ---
raster_coords        <- as.data.frame(prob_noisy_norm, xy = TRUE, na.rm = TRUE)
rainfall_pred        <- extract(rainfall,
                                 vect(raster_coords,
                                      geom = c("x", "y"),
                                      crs  = crs(rainfall)))[, 2]
ndvi_pred <- extract(ndvi,
                      vect(raster_coords,
                           geom = c("x", "y"),
                           crs  = crs(ndvi)))[, 2]



A_pred               <- inla.spde.make.A(mesh,
                                          loc = as.matrix(raster_coords[, c("x", "y")]))
spatial_mean         <- model$summary.random$spatial$mean

pred_vals <- model$summary.fixed["intercept", "mean"] +
             model$summary.fixed["rainfall",  "mean"] * rainfall_pred +
             model$summary.fixed["ndvi",      "mean"] * ndvi_pred +
             as.vector(A_pred %*% spatial_mean)

raster_coords$pred_intensity <- exp(pred_vals)      # LGCP intensity surface
raster_coords$pred_prob      <- plogis(pred_vals)   # occupancy probability

# Also extract posterior uncertainty (standard deviation)
spatial_sd           <- model$summary.random$spatial$sd
pred_sd              <- as.vector(A_pred %*% spatial_sd)
raster_coords$pred_sd <- pred_sd

# Convert to rasters
intensity_rast <- rast(raster_coords[, c("x", "y", "pred_intensity")],
                        crs = crs(prob_noisy_norm))
prob_rast      <- rast(raster_coords[, c("x", "y", "pred_prob")],
                        crs = crs(prob_noisy_norm))
sd_rast        <- rast(raster_coords[, c("x", "y", "pred_sd")],
                        crs = crs(prob_noisy_norm))

# Plot
par(mfrow = c(1, 3))
plot(intensity_rast, main = "Predicted intensity (LGCP)")
plot(prob_rast,      main = "Predicted occupancy probability")
plot(sd_rast,        main = "Posterior uncertainty (SD)")
```


